---
title: "Fuzzy difference and data primitives: a transparent approach for
supporting different definitions of `forest` in the context of
REDD+"
author: "Alexis Comber^1^, Werner Kuhn^2^"
output: 
  pdf_document: 
    keep_tex: yes
---


^1^ School of Geography, University of Leeds, Leeds, LS2 9JT, UK

email: a.comber@leeds.ac.uk


^2^ Department of Geography, University of California Santa Barbara, USA

email: werner@ucsb.edu

# Abstract

This paper explores the use of fuzzy difference methods in order to
understand the differences between forest classes. The context
for this work is provided by REDD+, which seeks to reduce the net
emissions of greenhouse gases by rewarding the conservation of forests
in developing countries. REDD+ requires that local inventories of forest
are undertaken and payments are made on the basis of the amount of
forest (and associated carbon storage). 
At the most basic level
this involves classifying land into forest and non-forest. However, the
critical issues affecting the uptake, buy-in and ultimately the success
of REDD+ are the lack of universally agreed definition of forest to
support REDD+ mapping activities, and where such a definition is
imposed, the marginalization of local community voices and local
landscape conceptualizations. This tension is at the heart of REDD+. The
methods described in this paper to address these issues link methods to quantify
fuzzy land cover change to notions of data primitives, which have been
previously proposed as an approach to move between land cover classes
with different semantics. Together these methods provide for transparent
translations between alternative conceptualizations of forest, allowing for
plural notions of forest to be mapped and quantified. In particular, they allow for moving from an object-based notion of forest (and land cover in general) to a field-based one, entirely avoiding the need for forest boundaries.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
cach.val = T
```
```{r functions, include=FALSE}
# A function for captioning and referencing figures
fig <- local({
    i <- 0
    ref <- list()
    list(
        cap=function(refName, text) {
            i <<- i + 1
            ref[[refName]] <<- i
            paste( text, sep="")
        },
        ref=function(refName) {
            ref[[refName]]
        })
})
```

# 1. Introduction

This paper suggests and applies an alternative approach to classifying and
mapping land cover, using the example of forest mapping in the context of
Reducing Emissions from Deforestation and forest Degradation (REDD and
REDD+). REDD+ initiatives seek to reduce the net emissions of greenhouse
gases by rewarding the conservation of forests in developing countries
(Angelsen, 2009). REDD+ requires that local inventories of forest are
undertaken and payments are made on the basis of the amount of forest
(and associated carbon) that are mapped. At the most basic level this
involves classifying land into forest and non-forest.

The problem encountered by many forest monitoring strategies
under REDD+ is how to accommodate divergent conceptualizations of \emph{forest} such that local interpretations are reflected in forest mapping while providing a transparent tool for carbon
accounting. The International Panel on Climate Change (IPCC) recommends
the use of an internationally recognized forest definition (e.g., that of the Food and
Agricultural Organisation, FAO) but even within the United Nations Framework Convention on Climate Change (UNFCCC)
there are no agreed definitions (Romijn et al, 2013). The flexibility of
REDD+ allows forest to be defined locally, but ultimately requires
mappings to be transparent (Gupta et al, 2014; Ryan et al., 2014), in order for
forest estimates to be \textit{accurate} and payments to be justified.
There are well established reasons for not having a standard definition of
forest (see Comber et al 2007, for an extensive discussion on the
problems caused by the imposition of standards in land cover mapping). Not least among these reasons is that local considerations and landscape
conceptualizations would be ignored (Hoeschele, 2000; Robbins 2001) to the
detriment of local stakeholders. However, there
can be considerable disagreement over the amount of forest, when mapped
under different classifications. For example, Romijn et al (2013)
compared the forest stock levels in Indonesia when mapped using local
forest definitions and FAO definitions and found the extent of forest to
be 27\% higher when using the national definition.

The critical issue addressed by this paper is how to allow for the divergence of landscape
conceptualizations, avoid
divergent and possibly inaccurate estimates of forest extent. The
paper proposes a method based on a mix of data primitives and soft
classifications that allow the critical dimensions associated with land
cover and land use to be captured \textbf{and} the inherent relative uncertainty
when any two mappings are compared. This approach provides a shared,
transparent, robust and well understood method for mapping and reporting
forested areas. It is transparent and
supports global policy initiatives such as REDD+, whilst at the same time
accommodating local divergence in forest semantics - and the way that
`forest' (and therefore the boundary with non-forest) is conceptualized
spatially.

Our analysis uses existing classifications of forest from the Global Land Cover (GLC) and from
FAO and shows how these can be re-interpreted using soft classifications
to explore the overlaps and boundaries between different versions of
forest. In this way the paper develops approaches that could be used in
the context of REDD+ initiatives to support multiple views of the same
landscape to support divergent objectives. Such transparency can
overcome some of the problems of exclusion and back door land reforms
that are commonly experienced during REDD+ initiatives and mapping
(Ezzine-de-Blas, 2011), as well as the issues highlighted by Romijn et
al (2013) of how to define `forest'. In our approach, one can define
forest however one wants and link it to another definition.

# 2. Background

To situate our approach, we first provide the necessary background on land cover classification, the problems associated with forest definitions, and the method of fuzzy boundaries and fuzzy differences. 

## 2.1 Land cover classification
Classification, the process of sorting real world phenomena into
categories or classes, is a core activity within geography and many
other disciplines. Classes describe groups of social and environmental
phenomena with broadly similar characteristics. Land cover classes are
used to describe the physical characteristics of the earth's surface and
land use classes describe the socio-economic activities thereon. Typical land
cover classes include water, grass, and forest; typical land use classes
include agriculture, forestry, and urban. There is an extensive
literature on the different philosophical underpinnings of the way that
land cover and land use classes are conceptualized (e.g.~Comber et al.,
2005; Fisher et al, 2005), also for specific classes such as forest
(Bennet, 2001, Comber et al, 2008). This application based literature is
complemented by more general considerations of the nature of
classification (e.g.~Lakoff, 1987) and by more formal research
considering the nature of how to represent objects, processes, and
relationships of geographic phenomena in computer models and analyses (e.g.~Kuhn 2005, 2012).

Classification typically allocates each item uniquely to one class and
this is true for land cover and land use, although much theoretical work
exists suggesting alternative, soft classifications  such as those supported by Fuzzy Sets (e.g.~Fisher 2010). All global land cover data produced for monitoring purposes and used as inputs to
global climate change models adopt a crisp classification with hard and crisp
boundaries between the resulting land cover objects of different classes. Thus, land cover classification is
a process of homogenization that ignores any within class variation and, more critically, excludes any consideration of the potential for
uncertainty in boundaries between classes. This contrasts with a
well developed literature on boundary uncertainty within geography (Smith, 1995;
Burrough and Frank 1996), reflecting a vernacular definition of geography as the art
of drawing lines on a map that do not exist in reality.

Classifications with crisp boundaries provide a convenient and
familiar framework for representing environmental and social processes
that are readily understood by policy makers and initiatives. For
example, in the sphere of climate change, REDD+ and climate models seek
to standardize or harmonize definitions of forest and develop crisp
mappings of forested areas that may not exhibit crisp edges in reality
and whose classification may be deeply
contested. This culture of class definition
standardization for crisp mapping creates baseline inventories that ignore
local variations in how land cover classes such as forest are
conceptualized, their semantics and the functions associated with them. This can have serious political implications
for the way that land is managed locally and for local political discourses
around land. This has become evident with the roll out of REDD+ in many developing
countries, where fixed IPCC / UN definitions of `forest`, for the
purpose of carbon accounting, are being used to drive land reform and
privatization agendas. Local land cover interpretations are excluded
from local and global decision making.

## 2.2 Forest definitions and REDD+

The REDD and REDD+ initiatives (Reducing Emissions from Deforestation
and forest Degradation), as set up by the United Nations Framework
Convention on Climate Change (UNFCCC) in 2005 and enhanced in 2007 to
stimulate actions related to conservation and sustainable forest
management (hence the `+`), aim at mitigating climate change by reducing
net emissions of greenhouse gases by financially rewarding enhanced
forest management in developing countries. In simple terms, poor
countries with forests are to be paid to \emph{not} cut down their
trees, with payments based on the amount of trees and carbon storage capacity they have,
although the precise details are still to be decided. In order to
support payments, the UN have suggested that countries should develop
robust and transparent forest monitoring system to record forest and
carbon stocks and changes using a combination of remote sensing and
field surveys (UNFCCC, 2009), with each country defining what
constitutes \emph{forest}. This was to overcome the deeply contested
nature of what is a `forest` (and land cover / land use classification
in general), well established in the literature (see for example
Hoeschele 2000; Bennet 2001; Robbins 2001; Smith and Mark 2001; Comber
et al, 2005) and reflects commendable courage by the remote sensing
community that is driving the REDD+ initiative. However, REDD+
activities require forest to be demarked, presenting a number of
challenges, two of them being of critical importance.

First, any particular definition of forest and the associated mapping will
determine the reported amount of forest changes, which will in turn
affect the estimates of carbon storage (Magdon and Kleinn, 2012; Romijn et al,
2013). There are hundreds of forest definitions as documented by Lund
(2015) and explored conceptually by Bennett (2001) and Comber (2005)
with profoundly different threshold parameters for height, stripwidth,
minimum area, species (some include grasses such as bamboo), canopy,
management / plantation cover etc. These are international, national and
sub-national and reflect locally important variations in the concept of
`forest` and its semantics - what the concept is in the local (not
just REDD+) context. In the context of REDD+, Morales-Barquero et al
(2014) highlight the difficulties in operationalizing local and national
level REDD+ projects and programmes in Mexico. They present a framework
to support local definitions and measurements of forest degradation that
support both local and national objectives and suggest the use of
comparative biophysical benchmarks for assessing degradation. More
recently, Chazdon et al (2016) examined historical forest concepts and
definitions and documented how these relate to variations in
socio-economic activity. They note a number of problems with top-down
forest definitions for assessing global changes in forest stocks: their failure to distinguish between natural / plantation forests and the lack of
consideration of the qualities and trajectories of forested areas in
standard approaches. In order to try to accommodate this divergence in
the concept of forest, and to support consistency in carbon accounting,
the UN requires robust and transparent national reporting of the
measurement systems, the remote sensing and ground based forest carbon
inventory data, the methods and the forest definition used in national
measurements (Herold et al., 2012a, 2012b). The impacts of such semantic
variations are well documented. Romijn et al (2013) compared measures of
deforestation using 3 definitions of forest in Indonesia - the FAO
definition, a `natural forest` definition and a national forest
definition - and found large differences between the amounts of
deforestation.

Second, national definitions and associated mappings are deeply
political exercises that are frequently used to drive secondary local
political agendas that exclude and are in conflict with local
communities. Mapped forests, however they are defined, are presented
locally as facts, which are then manipulated towards the interests of
the state, excluding community views, perceptions and opinions from the classification and measurement activity. For example, there are cases of REDD+
projects being used locally to support the violent eviction of people
from the land in many countries (Himmelfarb, 2012; Cavanagh and
Benjaminsen, 2014; Grainger and Geary, 2011; Nel and Hill, 2013; Lyons
and Westoby, 2014; Forest Peoples Programme, 2014; Beymer-Farris and
Bassett, 2012). In other cases, national definitions are imposed and the
subsequent mappings are used to support land privatization agendas and
forest commodification with the objective of removing commonly held land
from collective ownership (Ece, 2015). Critical to the success of
REDD+, there are also situations identifying the financial beneficiaries
of carbon payments because the forested land lacks natural ownership
divisions. A further negative dimension of REDD+ is that it promotes the
obfuscation and overlooking of the opinions and views of
forest-dependent communities and indigenous forest-dwelling populations
in developing countries. Local understanding and knowledge are important
for the forest conservation objectives of REDD+. For example, Bong et al.
(2016) compared community knowledge of deforestation and degradation
drivers amongst different Indonesian villages and found local knowledge
to be key to understanding the local impacts of deforestation
drivers. They highlighted the importance of incorporating local
knowledge and conceptualizations within definitions of forest to develop
more locally appropriate REDD+ monitoring systems. However, some
communities in likely REDD+ countries do not conceptualize forest at all
(Niclas Burenhult, pers comm) and where they do this has very nuanced
and spiritual meaning. Specifically, although the people of Jahai in the in the Malay Peninsula live in forests, they have no concept of forest (Burenhult, 2009). Instead it
is their home and the nearest terms they have for forest like things
describe leaves and trees as well as canopy and floor, covered area and
exposed area.

The conclusion is that the requirement of REDD+ to demarcate forest is
problematic in many places. Crisp mappings of forest do not reflect the many
cultural, linguistic and ownership aspects of forest, however the
category is defined. Initiatives like REDD+, however well
meaning they are, result in a clash of categories with the consequence that
important forest related parameters are overlooked. These wide ranging
negative implications for forest-dependent communities and indigenous
forest-dwelling populations threaten forest conservation in developing
countries and as a consequence have been referred to as the dark side of
REDD+ (Cavanagh et al., 2015).

## 2.3 Fuzzy Boundaries and difference 

Using fuzzy land classifications each piece of land (for example, the piece corresponding to a pixel in a satellite image) is
allocated a degree of membership to each class in the range {[}0,1{]}
(Fisher and Pathirana, 1990). This allows each piece of land to have partial
membership to more than one land cover class. For these reasons fuzzy approaches
are frequently referred to as \emph{soft} classifications as the fuzzy
outputs include and represent some of the inherent uncertainty
associated with allocating image elements (such as pixels) to classes
(Fisher, 1997). This is in contrast to \emph{hard} or Boolean
approaches, which do not accommodate any uncertainty. For these reasons
fuzzy sets and fuzzy classification algorithms have been suggested as
appropriate approaches in remote sensing analyses for representing land
cover objects which may or may not fit neatly into the classification
scheme and / or the sampling frame (pixel).

In a standard land cover classification, class summaries or centres are
created that describe the (proto-) typical class properties usually in
an \(n\)-dimensional feature space, for example representing different remote sensing image bands,
or as in this case data primitives, described below. In a Boolean classification each
piece of land is allocated to the class to which it is
nearest. Different algorithms create the class centres in different ways
and measure distance in \(n\)-dimensional space in different ways. The
issue that Fuzzy classifications seek to address is
that a piece of land may be \emph{near} to more than one class, suggesting
that it contains some of their properties and resulting in a degree of
ambiguity (Fisher et al., 2006a) in the allocation.

A second commonly arising problem that fuzzy classifications seek to
address relates to the sampling frame. The assumption in mapping land
cover using remote sensing is that the objects of interest on the
ground are adequately described and captured by the imagery. The
inference is that the spatial scale of the processes on the ground (such
as land cover) are larger that then the pixel and that contiguous pixels
will describe the extent of these processes. In reality there are many
sub-pixel objects on the ground (Fisher, 1997) arising from a mismatch
between the scale of observation and the granularity of the processes
being observed (Fisher, 2007). Fuzzy sets seeks to accommodate a number
of problems related to the pixel: sub-pixel objects (less than the size of a pixel), trans-pixel objects (linear objects crossing pixels), intergrades (where land cover types merge into each other over space) and boundaries (where multiple covers meet within one
pixel). 

There is a long standing body of research describing the application of
fuzzy sets to land cover and this has been extended by some researchers who
have developed methods for related analyses: fuzzy accuracy, fuzzy
change detection, Type-2 and Type-N fuzzy approaches (Fisher, 2010;
Fisher et al., 2007). Conceptually, fuzzy approaches have potential value
to operational land cover monitoring exercise such as REDD+ because they
can explicitly accommodate the uncertainty inherent to classification.
Here these the concepts of fuzzy Minimum Interval, fuzzy Bounded
Difference and fuzzy Loss and Gain are extended to examine the
differences between two fuzzy classifications as described in the next
section.

## 2.4 Land cover objects vs land cover fields
The land cover classification process described above results in crisp and hard boundaries between classes in attribute space as well as between the resulting regions on the ground. It is therefore tempting to interpret these regions as land cover objects, with boundaries and identity. To quantify a country's amount of, for example, forest, one simply adds the surface areas of all forest objects in that country at a given time. 

Yet, for the reasons stated in the previous subsection, such an outcome of land cover classification is not satisfactory. In particular, for the case of forests, it is unsuitable to deal with partial deforestation and forest degradation. Forests do not simply disappear "from their boundaries inward" by converting pieces of forests to non-forests. Rather, deforestation and forest degradation proceed as gradual reductions of \emph{forestness} of pieces of land. They may occur, for example, through a systematic thinning of tree density. 

It turns out that fuzzy boundaries and differences, or rather fuzzy membership values for pieces of land in a land use category, provide exactly the mathematical model needed to deal with such an idea of forestness. As the following analysis section will show, they generate feature vectors for each piece of land corresponding to, for example, a pixel. In Geographic Information Systems (GIS), such arrays of vectors of observed values are known as vector fields. Thus, using the classic dichotomy of fields vs objects (Couclelis 1992), we conclude that forests are much more adequately conceptualized as fields rather than objects. 

# 3. Analysis
o illustrate the proposed method, this paper uses the case study of translating
land cover data describing forest in North America to another
classification. In this case we convert the 6 GLC forest classes to the
FAO class of forest. The approach is to characterize the
classified data using data layers referred to as data primitives (Comber
et al., 2008; Wadsworth et al, 2008) which enable each class to be
described at the most fundamental level. In the terminology of GIS field models, data primitives are the dimensions of the forest field vector. Here we use data describing
vegetation height, photo-synthetic activity, soil wetness, human
disturbance and seasonality. The positions of each class in these
dimensions are extracted. Then a new class of forest is defined using
the data primitives, and subsequently mapped. The differences between
the two soft, fuzzy mappings of `forest' within the primitives are
explored and shown to provide a transparent and translatable comparison
of different definitions of forest.


## 3.1 Data
The Global Land Cover (GLC2000) data were downloaded and extracted for
the USA. Five datasets to provide measures that approximated to the
dimensions of vegetation height, photo-synthetic activity, soil wetness,
human disturbance and seasonality were obtained from the
following open data portals:

\begin{itemize}
\item
  \textbf{Vegetation height}: from \emph{Existing Vegetation Height}
  data from
  \url{http://www.landfire.gov/version_comparison.php?mosaic=Y}
\item
  \textbf{Biomass}:from
  \url{http://landcarbon.org/categories/biomass-c/download/} g
  Carbon/m2.
\item
  \textbf{Soil wetness / soil moisture}: from
  \url{http://giovanni.gsfc.nasa.gov/giovanni} for the month of June for
  0-100cm depth
\item
  \textbf{Human disturbance}: proximity to mapped Urban and Managed land
  from GLC (classes 16, 17, 18, 22)
\item
  \textbf{Seasonality}: phenology growing season data from the MODIS
  archive: Duration DUR Length of photosynthetic activity (number of
  days): \url{https://lta.cr.usgs.gov/emodis_phen}
\end{itemize}

These were all resampled to 0.0217 of a degree with the mean value used
to aggregate the data, and clipped to a smaller case study area. The
spatial distributions of these variables and the GLC2000 data are shown
in Figures 1 and 2.

```{r eval=T, echo=F, cache = F, message=F, warning=F, results="hide"}
library(GISTools)
library(raster)
library(plotrix)
library(knitr)
library(OpenStreetMap)
library(e1071)
library(rgdal)
library(frbs)
### load in the variable data
library(repmis)
source_data("https://github.com/lexcomber/ForestPaper/blob/master/forest_paper_gs.RData?raw=True")
source_data("https://github.com/lexcomber/ForestPaper/blob/master/forest_paper_hd.RData?raw=True")
source_data("https://github.com/lexcomber/ForestPaper/blob/master/forest_paper_pr.RData?raw=True")
source_data("https://github.com/lexcomber/ForestPaper/blob/master/forest_paper_sw.RData?raw=True")
source_data("https://github.com/lexcomber/ForestPaper/blob/master/forest_paper_vh.RData?raw=True")
source_data("https://github.com/lexcomber/ForestPaper/blob/master/forest_paper_glc.RData?raw=True")

ex <- extent(-87, -77, 32, 38) #82, 35
v2 <- crop(vh, ex)
p2 <- crop(pr, ex) 
s2 <- crop(sw, ex)
h2 <- crop(hd, ex)
g2 <- crop(gs, ex)
r <- stack(v2, p2, s2, h2, g2)
names(r) <- c("vh", "pr", "sw", "hd", "gs")
#r <- stack(vh, pr, sw, hd, gs)
# create df
df <- getValues(r)
colnames(df) <- c("vh", "pr", "sw", "hd", "gs")
index.df <- apply(df, 1, function(x) is.na(sum(x)))

p1 <- spplot(r, "vh", key.space = "right", pch=15, cex=0.6, 
            col.regions=brewer.pal(6,"Greens"), cuts=5,
            par.settings=list(fontsize=list(text=10)),
            main=list(label="Vegetation Height",cex=1.), asp = 1)
p2 <- spplot(r, "pr", key.space = "right", pch=15, cex=0.6, 
            col.regions=brewer.pal(6,"Reds"), cuts=5,
            par.settings=list(fontsize=list(text=10)),
            main=list(label="Biomass",cex=1.), asp = 1)
p3 <- spplot(r, "sw", key.space = "right", pch=15, cex=0.6, 
            col.regions=brewer.pal(6,"Blues"), cuts=5,
            par.settings=list(fontsize=list(text=10)),
            main=list(label="Soil Wetness",cex=1.), asp = 1)
p4 <- spplot(r, "hd", key.space = "right", pch=15, cex=0.6, 
            col.regions=brewer.pal(6,"Oranges"), cuts=5,
            par.settings=list(fontsize=list(text=10)),
            main=list(label="Lack of Human Disturbance",cex=1.), asp = 1)
p5 <- spplot(r, "gs", key.space = "right", pch=15, cex=0.6, 
            col.regions=brewer.pal(6,"YlOrRd"), cuts=5,
            par.settings=list(fontsize=list(text=10)),
            main=list(label="Seasonality",cex=1.), asp = 1)

glc <- as(glc, "RasterLayer")

```

```{r, fig.height=7.2, cache = F, echo = F, eval=T,  message=F, warning=F, fig.cap=fig$cap("", "The data primitives.")}
# 
print(p1, position = c(0,0,0.5,0.5), more = T)
print(p2, position = c(0.5,0,1,0.5), more = T)
print(p3, position = c(0.5,0,1,1.1), more = T)
print(p4, position = c(0,0,0.5,1.1), more = T)
print(p5, position = c(0,0,0.5,1.7), more = F)
```

```{r, fig.height=7.2, cache = cach.val, echo = F, eval=T,  message=F, warning=F, fig.cap=fig$cap("", "GLC2000 data: green and brown are forest classes, pink is agriculture and red is urban.")}
# load some data to generating training
cols <- rgb(labs[,3:5])
par(mar = c(0,0,1,0))
plot(as(glc, "SpatialPointsDataFrame"), 
	col = cols[getValues(glc)], 
	pch = 15, cex = 0.5)

```

## 3.2 Class centres

The typical characteristics of each of the GLC2000 classes in each of
the five dimensions were extracted and are shown in Table 1. This was done by selecting a sample of
GLC2000 pixels from each class, with the sample number weighted by the
distribution of the class in the study area. 


```{r eval=T, echo=F, cache = cach.val, message=F, warning=F, results="hide"}
# stratified sample from CLIPPED GLC data
# crop GLC
ex <- extent(glc)
ex[2,] <- -81.27207
glc2 <- crop(glc, ex)
glc.sp <- as(glc2, "SpatialPointsDataFrame")
proj4string(glc.sp) <- CRS("+proj=longlat")
# add new class - never actually used
ind <- match(glc.sp@data$band1, labs$VALUE)
glc.sp@data$band2 <- labs$Class[ind]
# sample classes
table(glc.sp$band1)
class.list <- as.numeric(names(which(table(glc.sp$band1) > 100)))
# some weights - never used
wt.list <- which(table(glc.sp$band1) > 100)
wt.list <- as.numeric(names(table(glc.sp$band1))[wt.list])
wt.list <- table(glc.sp$band1)[as.character(wt.list)]
wt.list <- wt.list/sum(wt.list)
# sample
sample.sp <- glc.sp[0,]
for (i in 1: length(class.list)) {
	class.i = class.list[i]
	ind.i <- glc.sp$band1 == class.i
	data.i <- glc.sp[ind.i, ] 
	set.seed(100)
	s.i <- sample(1:nrow(data.i), 100000 * wt.list[i])
	sample.sp <- spRbind(sample.sp, data.i[s.i,])
}
index <- match(class.list, rownames(table(glc.sp$band1)))
tab <- rbind(table(sample.sp$band1), table(glc.sp$band1)[index])
rownames(tab) <- c("sample", "total no")
tab <- t(tab)
rownames(tab)<- labs$CLASSNAMES[class.list]


# overlay sample with variable data  
v.s <- over(spTransform(sample.sp, 
	proj4string(as(r, "SpatialPixelsDataFrame"))),  
	as(r, "SpatialPixelsDataFrame"))
# rename classes 1:n for frbs
class.list <- sort(unique(sample.sp$band1))
new.class <- 1:length(class.list)
index <- match(sample.sp$band1, class.list)
sample.sp$Species <- new.class[index]
v.s <- cbind(v.s, sample.sp$Species)
names(v.s)[6] <-"Species"
# class centres (not rescaled) from sample
cents.mat <- matrix(ncol = 5, nrow = 0, data = 0)
colnames(cents.mat) <- names(v.s)[-6]
for (i in 1:length(class.list)) {
	i.i <- sample.sp$band1 == class.list[i]
	vs.i <- v.s[i.i,]
	med.i <-apply(vs.i, 2, function (x) median(x, na.rm = T))[1:5]
	cents.mat <- rbind(cents.mat, med.i)	
}
rownames(cents.mat) <- labs$CLASSNAMES[class.list]
```  
  


```{r eval=T, echo=F, cache = cach.val, message=F, warning=F, results="hide"}

# now do classification - with cmeans
index <- apply(df, 1, function(x) is.na(sum(x)))
cl <- cmeans(df[!index,], centers = cents.mat, m = 4.5, 
	rate.par = 0.97, iter.max = 500)
# put into raster
res <- matrix(nrow = length(index), ncol = length(class.list)+2)
clus <- cl$cluster
old.class <- class.list[clus]
res[!index, ] <- cbind(cl$membership, cbind(clus, old.class))
fdall <- brick(r, nl = length(class.list)+2) # r is already a brick!
for ( i in 1: (length(class.list)+2)) {
	fdall <- setValues(fdall, res[,i], layer = i)
	#cat(i, "\t")
}
#plot(fdall, 9, col = cols[class.list])
# ok so too muach water but this will do
# add ALL forest MF class
tmp <- rowSums(res[, 1:3])
fdall <- 	setValues(fdall, tmp, layer = 11)
names(fdall) <- c(paste("c",class.list, sep = ""), "clus", "PredGLC", "AllFor")

# now do new forest class
# smaple for that
set.seed(100)
samp.list <- sample(length(glc.sp), 10000)
sample2.sp <- glc.sp[0,]
for (i in 1:5000) {
	samp.i <- samp.list[i]
	data.i <- glc.sp[samp.i, ] 
	if (data.i$band1 < 7) sample2.sp <- spRbind(sample2.sp, data.i)
}
ol <- over(spTransform(sample2.sp, proj4string(as(r, "SpatialPixelsDataFrame"))),  
	as(r, "SpatialPixelsDataFrame"))
index <- which(ol$vh > 5 & (ol$pr > max(getValues(r$pr), na.rm = T)* 0.1))
#summary(index)
sample.sp2 <- sample2.sp[index,]
ol <- ol[index,]
tab3 <- round(apply(ol, 2, fivenum), 1)
set.seed(100)
samp.list <- sample(length(glc.sp), 10000)
sample3.sp <- glc.sp[0,]
for (i in 1:5000) {
	samp.i <- samp.list[i]
	data.i <- glc.sp[samp.i, ] 
	if (data.i$band1 > 6) sample3.sp <- spRbind(sample3.sp, data.i)
}
ol <- over(spTransform(sample3.sp, proj4string(as(r, "SpatialPixelsDataFrame"))),  
	as(r, "SpatialPixelsDataFrame"))
index <- which(ol$vh < 5)
#summary(index)
sample.sp3 <- sample3.sp[index,]
ol <- ol[index,]
tab2 <- round(apply(ol, 2, fivenum), 1)

new.cents <- rbind(tab3[3,], tab2[3,])
rownames(new.cents) <- c("FAO Forest", "FAO not-Forest")

index <- apply(df, 1, function(x) is.na(sum(x)))
cl2 <- cmeans(df[!index,], centers = new.cents, m = 4.5, 
	rate.par = 0.97, iter.max = 500)
# put into raster
res <- matrix(nrow = length(index), ncol = 3)
clus <- cl2$cluster
res[!index, ] <- cbind(cl2$membership, clus)
fdall.new <- brick(r, nl = 3) # r is already a brick!
for ( i in 1:3) {
	fdall.new <- setValues(fdall.new, res[,i], layer = i)
	#cat(i, "\t")
}

```

```{r eval=T, echo=F}
library(knitr)
kable(tab, booktabs = T, format = 'latex', align = 'l', digits = 0, caption = "The sample number and total distribution of the GLC2000 classes.")
```


```{r eval=T, echo=F}
library(knitr)
kable(round(cents.mat, 1), format = 'latex', booktabs = T, align = 'l', digits = 1, caption = "The median values for each GLC2000 in the 5 primitive dimensions: VH = Vegetation Height; PAR = Photosynthetic Activity; SW = Soil Wetness; LHD = Lack of Human Disturbance; GS = Seasonality.")
```

```{r eval=T, echo=F}
library(knitr)
kable(round(new.cents, 1), format = 'latex', booktabs = T, align = 'l', digits = 1, caption = "The median values for each FAO Forest classes in the 5 primitive dimensions.")
```


## 3.3 Fuzzy Classification: GLC2000 and FAO

A fuzzy c-Means classifier was used to generate soft classifications for
each pixel in the study area. This algorithm used was the implementation
of the c-Means fuzzy clustering method (e.g.~Bezdek, 1981) in the
\texttt{e1071} R package (Meyer et al., 2012). The classifier generates
fuzzy memberships in the range of {[}0, 1{]} to each class for each
pixel. In this case the class centres were those in Table 2 and the
classifier generated fuzzy memberships to the et of each class in the
five primitive dimensions. The fuzzy c-means algorithm seeks to to
minimise the objective function:

\begin{equation}
\sum_{i} \sum_{j} w_{i} u_{ij}^m d_{ij}
\end{equation}

where \(w_i\) is the weight of observation (pixel) \(i\), \(u_{ij}\) is
the membership of observation \(i\) in cluster (class) \(j\), and
\(d_{ij}\) is the Euclidean distance (dissimilarity) between observation
\(i\) and centre \(j\). The fuzzy cmeans algorithm generates a
membership to each class for each pixel, based on their closeness to the
class centres in Table 2.

The fuzzy memberships can be used to extract crisp values of the amount
of forest and non-forest, depending on the classes of forest that are
preferred, through the application of different alpha cuts. This in turn
allows the amount of carbon to be measured relative to the way that
\emph{forest} is defined.

Then by combining all of the memberships for the three forest classes, a
fuzzy class of forest from the GLC2000 classes was generated.

Finally, a second forest class was introduced. This was based on the he
FAO forest classes forest class definition
(\url{http://www.fao.org/docrep/006/ad665e/ad665e06.htm}) and included
the following key characteristics: tree crown cover, more than 10
percent canopy cover, minimum height of 5 metres although young and
`temporarily unstocked' (!) land can be included. By selecting a sample
of forest pixels with the characteristics of being greater than 5m and
have greater than 10\% canopy cover, a second set of class centres in
the five primitives was generated. These are shown in Table 3. A second fuzzy classification of
forest was then generated using the c-Means algorithm above, from the class
centres.

## 3.5 Fuzzy boundaries and difference
The method for quantifying fuzzy differences between 2 classification is
based on that suggested by Fisher et al (2006b). In this approach, the
land cover at any location is considered to have potential memberships
to each of the different classes being considered. That is, class
membership is considered as a fuzzy set. Fuzzy change is described using
the fuzzy change matrix which supports the calculation of fuzzy losses
and gains. This process is exemplified below with a walk-through
example.

The fuzzy confusion matrix, describing the intersection between the two
fuzzy forest classes can be derived using the minimum interval or using
the bounded difference of fuzzy loss and fuzzy gain. As Fisher (2006b)
noted the minimum interval is the standard approach in fuzzy sets but is
counter-intuitive when it is used to compare different classes -- it
only makes sense in the context of fuzzy land cover when comparing fuzzy
sets of the same class but captured at different times (i.e.~the
diagonal in a correspondence matrix). For these reasons a number of
alternative operators have been suggested (Klir and Yuan, 1995; Leung,
1988; Zadeh, 1965) and as Fisher et al (2006b, p166) note \emph{Of
these, the Bounded Difference between two fuzzy sets is the simplest
operator for which the results make the most sense}.

The notion of bounded difference underpins the fuzzy change matrix as
follows: the minimum interval is used to calculate the fuzzy changes
between fuzzy sets of the same class and the bounded difference is used
to determine all other fuzzy changes (i.e.~off-diagonal changes in
change matrix). Fuzzy loss and gain are derived from the marginal totals
of a change matrix describing transitions between land covers at Time 1
and Time 2. From the estimates of the amount (area) of fuzzy change
between each class can be obtained as well as the total amounts of loss
and gain.

### Minimum Interval
In a standard fuzzy sets approach the minimum operator is used to
determine the intersection between 2 fuzzy classifications, \(C_2\) and
\(C_2\):

\begin{equation}
\mu(C_2, C_2) = min(\mu(C_1, C_2))
\end{equation}

In the analyses below, this is used to determine the diagonal elements
in the fuzzy difference matrices.

### Bounded Difference
The bounded difference describes the fuzzy intersection of 2 land cover
types. It provides a conservative estimate of change because of its
logic and mathematics: essentially if the sum of fuzzy memberships at
any location are less than 1 then the membership to the intersection --
i.e.~fuzzy change -- will be zero. The bounded difference between 2
fuzzy sets A and B is defined as:

\begin{equation}
\mu(A\cap{B}) = max(0, \mu(A)+\mu(B)-1)
\end{equation}

### Loss and Gain 
In the methods suggested by Fisher et al (2006) the bounded difference
(Equation 3) is used to quantify the off-diagonal loss and gain at each
location (pixel) from the marginal totals of the fuzzy change matrix.


### Worked example
Consider 2 fuzzy classifications at 2 different classifications (\(C_1\)
and \(C_2\)) of 3 classes each (\(a\), \(b\) and \(c\)) on a 4 by 4
pixel grid, as described in Table 3. The diagonals on the fuzzy change
matrix are populated by the sum of the minimum intersections and the off
diagonals by the sums of the bounded difference.

```{r , echo = F, cache = F}
a1 <- matrix(c(0.50, 0.50, 0.43, 0.33, 
               0.50, 0.38, 0.29, 0.33, 
               0.20, 0.20, 0.40, 0.40, 
               0.20, 0.17, 0.40, 0.40), ncol = 4, byrow = T)
a2 <- matrix(c(0.50, 0.50, 0.25, 0.25, 
              0.38, 0.38, 0.25, 0.25, 
              0.14, 0.14, 0.40, 0.40, 
              0.14, 0.14, 0.25, 0.40), ncol = 4, byrow = T)
b1 <- matrix(c(0.33, 0.33, 0.14, 0.17, 
               0.33, 0.25, 0.29, 0.17, 
               0.60, 0.60, 0.20, 0.20, 
               0.60, 0.50, 0.20, 0.20), ncol = 4, byrow = T)
b2 <- matrix(c(0.33, 0.33, 0.50, 0.50, 
               0.25, 0.25, 0.50, 0.50, 
               0.43, 0.43, 0.20, 0.20, 
               0.43, 0.43, 0.38, 0.20), ncol = 4, byrow = T)
c1 <- matrix(c(0.17, 0.17, 0.43, 0.50, 
               0.17, 0.38, 0.43, 0.50, 
               0.20, 0.20, 0.40, 0.40, 
               0.20, 0.33, 0.40, 0.40), ncol = 4, byrow = T)
c2 <- matrix(c(0.17, 0.17, 0.25, 0.25, 
               0.38, 0.38, 0.25, 0.25, 
               0.43, 0.43, 0.40, 0.40, 
               0.43, 0.43, 0.38, 0.40), ncol = 4, byrow = T)

a <- cbind(a1, cbind(b1, c1))
b <- cbind(b2, cbind(b2, c2))

tab1 <- a
tab1 <- data.frame(tab1)
colnames(tab1) <- c("a1","","","", "b1","","","", "c1","","","")
tab2 <- a
tab2 <- data.frame(tab2)
colnames(tab2) <- c("a2","","","", "b2","","","", "c2","","","")
```


```{r eval=T, echo=F, cache = F}
library(knitr)
kable(tab1, format = 'latex', booktabs = T, align = 'l', digits = 2, caption = "Example fuzzy memberships for a hypothetical classification over 4 by 4 pixel dataste with 3 classes (a1, b1 and c1) ")

kable(tab1, format = 'latex', booktabs = T, align = 'l', digits = 2, caption = "Example fuzzy memberships for a second hypothetical classification, again with 3 classes (a2, b2 and c2)")

```

To illustrate this, the table of minimum interval for class \(a\) in
classes \(a1\) and \(a2\) is shown in Table 5, along with the bounded
difference between classes \(b1\) and \(c1\). The derived fuzzy
correspondence matrix is shown in Table 6.

```{r , echo = F, cache = F}

## some functions from 
# back_script_functions.R - JNCC work
# min interval
min.int <- function(c1t1, c1t2) {
	m.int <- matrix(0,nrow = dim(c1t1)[1], ncol = dim(c1t1)[2]) 
	for (i in 1:dim(c1t1)[1]) {
		for (j in 1:dim(c1t1)[2]) {
			m.int[i,j] <- min(c1t1[i,j], c1t2[i,j])
		}
	}
	return(m.int)}
## Bounded Difference 
b.dif <- function(c1t1, c1t2) {
	diff = matrix(0,nrow = dim(c1t1)[1], ncol = dim(c1t1)[2]) 
	for (i in 1:dim(c1t1)[1]) {
		for (j in 1:dim(c1t1)[2]) {
			#diff[i,j] <- max(0, (c1t1[i,j] - c1t2[i,j]))
			diff[i,j] <- max(0, (c1t1[i,j] + c1t2[i,j] -1))
		}
	}
return(diff)}
# fuzzy bounded difference CM
fuzzy.bd.cm <- function(t1, t2, class.names) {
	nclasses <- length(class.names)
	fcm.bd <- matrix(0,nrow = nclasses, ncol = nclasses)
	rownames(fcm.bd) <- class.names
	colnames(fcm.bd) <- class.names
	for (i in 1:nclasses) { 		#rows T1
		for (j in 1:nclasses) {	#cols T2
			if(i == j) fcm.bd[i,j] = sum(min.int(as.matrix(get(t1[i])), as.matrix(get(t2[j]))), na.rm = T)
			if(i != j) fcm.bd[i,j] = sum(b.dif(as.matrix(get(t1[i])), as.matrix(get(t2[j]))), na.rm = T)
		}
	}
	return(fcm.bd)
}

# table 5: min int and bouded different 
tab <- min.int(a1,a2)
tab2 <- b.dif(b1,c2)
tab <- cbind(tab, cbind(rep("", nrow(tab))), tab2)
colnames(tab) <- c("a1-a2", rep("", 4), "b1-c2", rep("", 3))
# table 6: fuzzy difference / change matrix
t1 <- c("a1", "b1", "c1")
t2 <- c("a2", "b2", "c2")
class.names <- c("a", "b", "c")
fcm <- fuzzy.bd.cm(t1,t2, class.names)
gain <- colSums(fcm)-diag(fcm)
loss <- rowSums(fcm)-diag(fcm)
fcm <- cbind(fcm, loss, gain)

```

```{r eval=T, echo=F, cache = cach.val}
library(knitr)
kable(tab, format = 'latex', booktabs = T, align = 'l', digits = 2, caption = "The minimum interval for class $a$ and the bounded difference between classes $b1$  and $c1$.")

kable(fcm, format = 'latex', booktabs = T, align = 'l', digits = 2, caption = "The fuzzy difference matrix with loss and gain for the worked example.")
```


# 4. Results

## 4.1 Fuzzy Forest

The fuzzy memberships to the three forest classes, whose class centres are
listed in Table 2, are shown in Figure 3, along with a combined
membership for all GLC2000 forest classes. The class centres for the
forest class arising from FAO class definition are shown in shown Table
8. The result of the fuzzy C-Means classification is shown in Figure 4.

```{r, fig.width=7.2, cache = cach.val, echo = F, eval=T,  message=F, warning=F, fig.cap=fig$cap("", "The fuzzy membership to the 3 GLC2000 forest classes in study area, and their combined fuzzy memberships to an overall forest class.")}
## Pass them in to your call to plot
arg <- list(at=1:length(class.list))

# plot forest classes (section 3.2)
plot(fdall[[c(1:3, 11)]],breaks= seq(0,1, 0.05), axis.args=arg, 
	col = colorRampPalette(brewer.pal(9,'GnBu'))(20),
	asp = 1, 
	main= c(gsub("Tree Cover, ", "", labs[class.list[1:3],2]), "overall"),
	legend.args=list(text='mf', side=4, font=2, line=0.5, cex=0.8), 
	legend.only=F, legend.width=1, legend.shrink=0.6, 
	axes = F, box = F, nc = 2)
```

 
```{r eval=T, echo=F}
library(knitr)
kable(round(new.cents, 1), format = 'latex', booktabs = T, align = 'l', digits = 1, caption = "The median values for the FAO Forest and not-Forest classes in the 5 primitive dimensions.")
```

```{r, fig.width=7.2, fig.height=3,cache = cach.val, echo = F, eval=T,  message=F, warning=F, fig.cap=fig$cap("", "The fuzzy membership to the the FAO Forest and not-Forest classes in study area.")}
## Pass them in to your call to plot
arg <- list(at=1:length(class.list))

# plot forest classes (section 3.2)
plot(fdall.new[[c(1:2)]],breaks= seq(0,1, 0.05), axis.args=arg, 
	col = colorRampPalette(brewer.pal(9,'YlOrRd'))(20),
	asp = 1, 
	main= c("FAO Forest", "FAO not-Forest"),
	legend.args=list(text='mf', side=4, font=2, line=0.5, cex=0.8), 
	legend.only=F, legend.width=1, legend.shrink=0.3, 
	axes = F, box = F, nc = 2)

```

## 4.2 Boundaries

Having generated the fuzzy classifications, the resulting boundaries between the two
forest classes (GLC2000 and FAO) can now be compared. It is
instructive to consider a sub-region in greater detail, in this case an
area to the North and West of Atlanta, Georgia. Figure 5 shows the
GLC-derived and FAO-derived fuzzy forest distributions, with greater
detail in Figure 6. It is evident that the FAO membership values are greater in
many places than the GLC memberships and this is reflected in the
summary statistics in Table 8. Figure 5 shows that, broadly, the presence
of forest is indicated in the same places (erroneously or not!) by both
definitions of forest, with varying degrees of membership, suggesting
different intensities of forest. The detail in Figure 6
illustrates the local variation in those membership intensities.

```{r eval=T, echo=F, cache = cach.val, message=F, warning=F, results="hide"}
library(tidyverse)
library(raster)
fd.c <- brick(fdall[[11]], fdall.new[[1]], ( ( (fdall[[11]]- fdall.new[[1]])) ))
names(fd.c) <- c("GLCFor", "FAOFor", "Diff")

tab2 <- summary(fd.c)[-6,1:2]
tab2 <- rbind(tab2, colSums(getValues(fd.c[[1:2]]), na.rm = T))
tab2 <- round(tab2,3)
colnames(tab2) <- c("GLC derived Forest", "FAO derived Forest")
rownames(tab2)[6] <- "Total"
```
 
```{r eval=T, echo=F}
kable(tab2, booktabs = T, align = 'l', digits = 2, caption = "The summaries of the Fuzzy memberships GLC and FAO derived Forest classes.")
```

```{r, fig.width=7.2, fig.height=4,cache = cach.val, echo = F, eval=T,  message=F, warning=F, fig.cap=fig$cap("", "Detail around Atlanta, Georgia of the differences in the fuzzy memberships to the the GLC (red) and FAO (blue) forest classes, with a map of difference, with yellow indicates similar fuzzy membership values, red where GLC fuzzy forest memberships are greater and Blue where FAO are greater.")}
# now lets start to examine the data detail
library(raster)
library(OpenStreetMap)
library(GISTools)
ex <- extent(c(-84.71745, -82.63263, 33.70063, 35.24159))
fdcc <- crop(fd.c, ex)
par(mar = c(0,0,0,0))
par(mfrow = c(1,3))
#plot(MyMap)
plot(projectRaster(fdcc[[1]], crs=osm()), 
	col = brewer.pal(9, "Reds"), axes = F, box = F, legend = F, 
	breaks= seq(.1,1, 0.1))
plot(projectRaster(fdcc[[2]], crs=osm()), 
	col = brewer.pal(9, "Blues"), axes = F, box = F, legend = F,
	breaks= seq(.1,1, 0.1))
plot(projectRaster(fdcc[[3]], crs=osm()), 
	col = brewer.pal(8, "Spectral"), axes = F, box = F, legend = F,
	breaks= seq(-0.6,0.2, 0.1))
```

```{r, fig.width=7.2, fig.height=3.7,cache = cach.val, echo = F, eval=T,  message=F, warning=F, fig.cap=fig$cap("", "The boundary conditions between Forest and not-Forest arising from the differently defined forest classes, with context from an Open Street Map backdrop. The fuzzy GLC forest is mapped in red and the FAO fuzzy forest class is mapped in blue, both with a transparency term.")}

ex <- extent(c(-84.6, -84.1, 33.9, 34.5 ))
fdcc <- crop(fd.c, ex)
ul <- as.vector(cbind(bbox(fdcc)[2,2], 
  bbox(fdcc)[1,1]))
lr <- as.vector(cbind(bbox(fdcc)[2,1], 
  bbox(fdcc)[1,2]))
# download the map tile
MyMap <- openmap(ul,lr,9)
# now plot the layer and the backdrop
par(mar = c(0,0,0,0))
par(mfrow = c(1,3))
plot(MyMap)
plot(MyMap)
plot(projectRaster(fdcc[[1]], crs=osm()), 
	col = add.alpha(brewer.pal(9, "Reds"), 0.5), axes = F, box = F, legend = F, add = T, 
	breaks= seq(.1,1, 0.1))
plot(MyMap)
plot(projectRaster(fdcc[[2]], crs=osm()), 
	col = add.alpha(brewer.pal(9, "Blues"), 0.5), axes = F, box = F, legend = F, add = T, 
	breaks= seq(.1,1, 0.1))
```

\newpage

## 4.3 Fuzzy difference
The classic approach to examining the intersection between two fuzzy sets
is to use fuzzy bounded difference, with explanatory fuzzy losses and
gains (as in Table 6). The losses and gains are evident in
the off diagonals of the fuzzy bounded difference change matrix. The
data for per pixel fuzzy memberships for the FAO and GLC classifications
of \texttt{forest} and \texttt{not-forest} were used the input to the
fuzzy difference analysis (Equation 3). The results are shown in Table 9
and quantify the similarities and differences in the membership to the two
classes of forest in this study. This provides a tractable and
transparent method to move between two classifications, regardless of
their semantics. The off diagonals in Table 6 provide the equivalent of
the losses and gains, relative to the forest class under consideration
and describe how the semantic boundaries between the two classes
intersect.

```{r eval=T, echo=F, cache = cach.val, message=F, warning=F, results="hide"}
library(raster)
# table 6?: fuzzy difference / change matrix
GLCfor <- as.matrix(fd.c[[1]])
GLCNfor <- 1-GLCfor
FAOfor <- as.matrix(fd.c[[2]])
FAONfor <- 1-FAOfor
class.names <- c("Forest", "Not-Forest")
g1 <- c("GLCfor", "GLCNfor")
f1 <- c("FAOfor", "FAONfor")
fcm <- fuzzy.bd.cm(g1,f1, class.names)
#gain <- colSums(fcm)-diag(fcm)
#loss <- rowSums(fcm)-diag(fcm)
#fcm <- cbind(fcm, loss, gain)
```

```{r eval=T, echo=F}
kable(fcm, format = 'latex', booktabs = T, align = 'l', digits = 2, caption = "The fuzzy difference matrix for the GLC (rows) and FAO (columns) classifications of forest.")
```

# 5. Discussuon and Conclusion
The REDD+ programme aims to reduce carbon emissions from deforestation
and forest degradation. A key tenet of the programme is to financially reward less developed countries for their preservation (or increase) of forest stocks. However, any mapping, including land cover mapping, is only a representation of some view of 
%I added "some view of"
reality (Comber et al., 2005) and different views of reality
exist between and within different stakeholders and communities, whether
they be citizens, scientists or involved in policy making (see Harvey
and Chrisman (1998) for the classic example of the negotiation of the
differences in land cover definitions).

The problems for REDD+ are two-fold. First, differences in definitions
of forest influence the amount of forest, and therefore carbon storage, that is
mapped. This undermines the ambitions of REDD+. For example, Romijn et
al (2013) found that different definitions of forest resulted in
different levels of Indonesian deforestation being identified. They
critically noted that any definition of forests that is not adapted to
national circumstances could lead to large areas of deforestation being
excluded and not accounted for under the REDD+ scheme, confusing the
carbon accounting. Other research has identified similar problems (Van
Noordwijk and Minan 2009; Magdon 2014; Gou, 2016). To better understand
the implications and origins of this confusion, some research has
sought to unpick forest classifications and how they relate to different
management objectives (Chazdon et al 2016). Although these authors do not
refer to semantics, their study clearly shows how different forest
semantics are driven by initiatives (such as the IPCC, biodiversity
conventions, or agroforestry fora), noting that the presence of different management
objectives `provides a perspective from which specific definitions are
created' (p.539). Second, REDD+ forest mapping activities are being used
to impose a specific view of the landscape, often at the expense of
communities who live and work in forested areas (eg Bong et al., 2016).
These communities, although key stakeholders, are frequently
excluded and marginalized from the REDD+ mapping process, receiving
little or none of the financial rewards, and often suffering from opportunistic
land reforms arising on the back of REDD+. There is plenty of evidence
of this situation in the literature from all over the world (eg Gizachew
et al 2017, Astuti and McGregor 2016; Larson et al, 2013; Sunderlin et
al, 2014; Lyster 2011; Ezzine-de-Blas et al 2011; Duchelle et al, 2014)

In this paper we have shown how data primitives and and fuzzy intersections can be used to quantify the differences between stocks of
forest when different forest classifications are applied. The data
primitive approach defines each forest class using data of soil wetness,
vegetation height, biomass, human disturbance and seasonality. These
provide a common set of orthogonal data layers (GIS field layers) and are fully described in
Comber (2008) and Wadsworth et al (2008). Fuzzy intersection methods
(Fisher, 2006b) were used to quantify the overlaps between the classes,
specifically fuzzy bounded difference.

Thus, the problem of divergence in classification, resulting from diverse
conceptualizations of forest (as identified by many authors in the
context of data integration (eg Bennet 2001, Comber et al 2005), can be
overcome, not through standardization (see Comber et al, 2007) as argued
by many in the policy arena and the remote sensing community, but
by allowing \emph{divergent} forest classifications and mappings. Their relationships can be
handled through commonly used and accepted uncertainty handling methods. Crucially, such an approach to forest mapping does not rely on any boundaries, be they crisp or vague. It only requires to be able to characterize for each piece of land considered the values of the dimensions (data primitives) in the forestness vector field. 


# Acknowledgements
The authors would like to thank Ross Purves and his team for their
invitation to take part in a workshop in 2016, which led to the further
development of these ideas. Some of the ideas in this work were
developed under a Joint Nature Conservancy Committee project \emph{Fuzzy
approaches for Developing and Evaluating Earth Observation-enabled
ecological land cover time series system} (JNCC Reference:
C12-0171-0589) and Lex Comber would like to thank the late and great
Prof Pete Fisher for some of the ideas in this paper: these were lifted
from work that he and Pete never managed to finish. Werner Kuhn acknowledges support from the University of California, Santa Barbara, for the Center of Spatial Studies. All of the analyses
and maps in this paper were undertaken in R, the open source statistics
software. The code and data used to construct this analysis as well as
the \texttt{RMarkdown} file used to create this paper are freely
available at \url{https://github.com/lexcomber/ForestPaper} - have a
play!


# References

Angelsen, A. (Ed.). (2009). Realising REDD+: National strategy and
policy options. CIFOR.\\

Astuti, R., \& McGregor, A. (2016). Indigenous land claims or green
grabs? Inclusions and exclusions within forest carbon politics in
Indonesia. The Journal of Peasant Studies, 1-22.\\

Bennett B, 2001, What is a forest? On the vagueness of certain
geographic concepts, Topoi 20 , 189-201\\

Beymer-Farris, B.A., Bassett, T.J., 2012. The REDD menace: resurgent
protectionism in Tanzania's mangrove forests. Glob. Environ. Change 22
(2), 332--341.\\

Bezdek, J.C., (1981). Pattern Recognition with Fuzzy Objective Function
Algorithms. New York: Plenum Press.\\

Bong, I. W., Felker, M. E., \& Maryudi, A. (2016). How Are Local People
Driving and Affected by Forest Cover Change? Opportunities for Local
Participation in REDD+ Measurement, Reporting and Verification. PloS
one, 11(11), e0145330.\\

Burenhult, N. (2009). Landscape terms and toponyms in Jahai: a field
report. Working Papers in Linguistics, 51, 17-29.\\

Burrough P A, Frank A U (Eds), 1996 Geographic Objects with
Indeterminate Boundaries (Taylor and Francis, London)

Cavanagh, C., Benjaminsen, T.A., 2014. Virtual nature, violent
accumulation: the `spectacular failure' of carbon offsetting at a
Ugandan National Park. Geoforum 56, 55-65.\\

Cavanagh, C. J., Vedeld, P. O., \& Trdal, L. T. (2015). Securitizing
REDD+? Problematizing the emerging illegal timber trade and forest
carbon interface in East Africa. Geoforum, 60, 72-82.\\

Chazdon, Robin L., Pedro HS Brancalion, Lars Laestadius, Aoife
Bennett-Curry, Kathleen Buckingham, Chetan Kumar, Julian Moll-Rocek, Ima
Clia Guimaraes Vieira, and Sarah Jane Wilson. ``When is a forest a
forest? Forest concepts and definitions in the era of forest and
landscape restoration.'' Ambio 45, no. 5 (2016): 538-550.\\

Comber, A.J., Fisher, P.F., Wadsworth, R.A., (2005). What is land cover?
Environment and Planning B: Planning and Design, 32:199-209.\\

Comber, A., (2008). The separation of land cover from land use with data
primitives. Journal of Land use Science, 3(4): 215--229\\

Comber, A.J., Fisher, P.F., Wadsworth, R.A., (2007). Land cover: to
standardise or not to standardise? Comment on `Evolving standards in
land cover characterization' by Herold et al. Journal of Land Use
Science, 2(4): 287--291\\

Couclelis, H. (1992). People manipulate objects (but cultivate fields): Beyond the raster-vector debate in GIS. In A. U. Frank, I. Campari, U. Formentini (Eds.), Theories and methods of spatio-temporal reasoning in geographic space. Berlin: Springer-Verlag: 6577\\

Duchelle, Amy E., Marina Cromberg, Maria Fernanda Gebara, Raissa Guerra,
Tadeu Melo, Anne Larson, Peter Cronkleton et al. ``Linking forest tenure
reform, environmental compliance, and incentives: lessons from REDD+
initiatives in the Brazilian Amazon.'' World Development 55 (2014):
53-67.\\

Ece, M. (2015). Representation through Privatisation. Responsive Forest
Governance Initiative, working paper 23 -- available from
\url{https://sdep.earth.illinois.edu/files/RFGI_Working_Papers/RFGI\%2023_Melis\%20Ece.pdf}\\

Ezzine-de-Blas, D., Brner, J., Violato-Espada, A. L., Nascimento, N.,
\& Piketty, M. G. (2011). Forest loss and management in land reform
settlements: Implications for REDD governance in the Brazilian Amazon.
Environmental Science \& Policy, 14(2), 188-200. \\

Fisher, P.F. (1997). The pixel: a snare and a delusion. International
Journal of Remote Sensing 18 (3), 679-685. \\

Fisher, P. F., and Pathirana, S. (1990). The evaluation of fuzzy
membership of land cover classes in the suburban zone. Remote Sensing of
the Environment, 34, 121-132. \\

Fisher, P., Comber, A., and Wadsworth, R., (2006a).~ Approaches to
uncertainty in spatial data.~ In Fundamentals of Spatial Data Quality,
edited by R.Devillers and R.Jeansoulin, ISTE, London. pp43-59 \\

Fisher, P., Arnot, C., Wadsworth, R., and Wellens, J. (2006b). Detecting
change in vague interpretations of landscapes. Ecological Informatics,
1, 163-178. \\

Fisher, P., Cheng, T., \& Wood, J. (2007). Higher order vagueness in
geographical information: empirical geographical population of type n
fuzzy sets. Geoinformatica, 11(3), 311-330. \\

Fisher, P. F. (2010). Remote sensing of land cover classes as type 2
fuzzy sets. Remote Sensing of Environment, 114(2), 309-321. \\

Fisher, P.F., Comber, A.J., Wadsworth, R.A., (2005). Land use and Land
cover: Contradiction or Complement. Pp. 85-98 in Re-Presenting GIS,
(eds. Peter Fisher, David Unwin), Wiley, Chichester. \\

Forest Peoples Programme, 2014. Annual Report. Available form
\url{http://www.forestpeoples.org/tags/annual-reports/publication/2015/annual-report-2014} \\

Gizachew, B., Astrup, R., Vedeld, P., Zahabu, E. M., \& Duguma, L. A.
(2017, March). REDD+ in Africa: contexts and challenges. In Natural
Resources Forum. Blackwell Publishing Ltd. \\

Gou, Y. (2016). The Role of Maps in Capturing Distal Drivers of
Deforestation and Degradation: A Case Study in Central Mozambique. In
Land Use Competition (pp.~91-109). Springer International Publishing. \\

Grainger, M., \& Geary, K. (2011). The New Forests Company and its
Uganda plantations. Washington DC, USA: OXFAM International.

Gupta, A., Vijge, M. J., Turnhout, E., \& Pistorius, T. (2014). Making
REDD+ transparent: The politics of measuring, reporting, and
verification systems. Transparency in Global Environmental Governance:
Critical Perspectives, 181. \\

Harvey F, Chrisman N, 1998, ``Boundary objects and the social
construction of GIS technology'' Environment and Planning A 30 1683 \-{}
1694 \\

Herold, M., Verchot, L., Angelsen, A., Maniatis, D., Bauch, S., 2012a. A
Step-wise Framework for Setting REDD+ Forest Reference Emission Levels
and Forest Reference Levels. CIFOR Infobrief, no. 52, April 2012. \\

Herold, M., Angelsen, A., Verchot, L.V., Wijaya, A., Ainembabazi, J.H.,
2012b. A stepwise framework for developing REDD+ reference levels. In:
Angelsen, A., Brockhaus, M., Sunderlin, W.D., Sunderlin, L.V. (Eds.),

Analysing REDD+; Challenges and Choices. CIFOR, Bogor, Indonesia,
pp.~279--299. \\

Himmelfarb, D., 2012. In the Aftermath of Displacement: A Political
Ecology of Dispossession, Transformation, and Conflict on Mt. Elgon,
Uganda. PhD Dissertation. University of Georgia, Athens. \\

Hoeschele W, 2000, ``Geographic information engineering and social
ground truth in Attappadi, Kerala State, India'' Annals of the
Association of American Geographers 90 293 \-{} 321 \\

Klir, G.J., Yuan, B., (1995). Fuzzy Sets and Fuzzy Logic: Theory and
Applications. Prentice Hall, Englewood Cliff. \\

Kuhn, W. (2005). Geospatial semantics: why, of what, and how?. In
Journal on data semantics III (pp.~1-24). Springer Berlin Heidelberg. \\

Lakoff, G. (2008). Women, fire, and dangerous things. University of
Chicago press. \\

Larson, A.M., Brockhaus, M., Sunderlin, W.D., Duchelle, A., Babon, A.,
Dokken, T., Pham, T.T., Resosudarmo, I.A.P., Selaya, G., Awono, A. and
Huynh, T.B., 2013. Land tenure and REDD+: The good, the bad and the
ugly. Global Environmental Change, 23(3), pp.678-689. \\

Leung, Y.C., (1988). Spatial Analysis and Planning under Imprecision.
Elsevier, New York. \\

Lund H. Gyde (2015). Definitions of Forest, Deforestation,
Afforestation, and Reforestation.
\url{http://home.comcast.net/~gyde/DEFpaper.htm} \\

Lyons, K., Westoby, P., 2014. Carbon colonialism and the new land grab:
plantation forestry in Uganda and its livelihood impacts. J. Rural Stud.
36, 13--21. \\

Lyster, R. (2011). REDD+, transparency, participation and resource
rights: the role of law. Environmental science \& policy, 14(2),
118-126. \\

Magdon, P., Kleinn, C., 2012. Uncertainties of forest area estimates
caused by the minimum crown cover criterion - a scale issue relevant to
forest cover monitoring. Environmental Monitoring and Assessment, 1--16. \\

Magdon, P., Fischer, C., Fuchs, H., \& Kleinn, C. (2014). Translating
criteria of international forest definitions into remote sensing image
analysis. Remote Sensing of Environment, 149, 252-262. \\

Meyer, D., Dimitriadou, E., Hornik, K., Weingessel, A., eisch, F., 2012.
e1071: Misc Functions of the Department of Statistics (e1071), TU Wien,
2012. R package version, pp.~1--6. \\

Morales-Barquero, L., Skutsch, M., Jardel-Pelez, E. J., Ghilardi, A.,
Kleinn, C., \& Healey, J. R. (2014). Operationalizing the definition of
forest degradation for REDD+, with application to Mexico. Forests, 5(7),
1653-1681. \\

Nel, A., Hill, D., 2013. Constructing walls of carbon--the complexities
of community, carbon sequestration and protected areas in Uganda. J.
Contemp. Afr. Stud. 31 (3), 421--440. \\

Robbins P, 2001, ``Fixed categories in a portable landscape: the causes
and consequences of land-cover categorization'' Environment and Planning
A 33 161 \-{} 179 \\

Romijn, E., Ainembabazi, J. H., Wijaya, A., Herold, M., Angelsen, A.,
Verchot, L., \& Murdiyarso, D. (2013). Exploring different forest
definitions and their impact on developing REDD+ reference emission
levels: A case study for Indonesia. Environmental Science \& Policy, 33,
246-259. \\

Ryan, C. M., Berry, N. J., \& Joshi, N. (2014). Quantifying the causes
of deforestation and degradation and creating transparent REDD+
baselines: a method and case study from central Mozambique. Applied
Geography, 53, 45-54. \\

Smith B, Mark D, 2001, ``Geographical categories: an ontological
investigation'' International Journal of Geographical Information
Science 15 591 \-{} 612 \\

Smith B, 1995, ``On drawing lines on a map'', in Lecture Notes in
Computer Science 988 Spatial Information Theory Eds A U Frank, W Kuhn
(Springer, Berlin) pp 475 \-{} 484 \\

Sunderlin, W. D., Larson, A. M., Duchelle, A. E., Resosudarmo, I. A. P.,
Huynh, T. B., Awono, A., \& Dokken, T. (2014). How are REDD+ proponents
addressing tenure problems? Evidence from Brazil, Cameroon, Tanzania,
Indonesia, and Vietnam. World Development, 55, 37-52. \\

UNFCCC, 2009. Methodological guidance for activities relating to
reducing emissions from deforestation and forest degradation and the
role of conservation, sustainable management of forests and enhancement
of forest carbon stocks in developing countries. Decision COP 15/4.
2009. \\

Van Noordwijk, Meine, and Peter Minang. If we cannot define it, we
cannot save it: forest definitions and REDD. ASB PolicyBrief 15 (2009).

Wadsworth, R., Balzter, H., Gerard, F., George, C., Comber, A. and
Fisher, P. (2008). An environmental assessment of land cover and land
use change in Central Siberia using Quantified Conceptual Overlaps to
reconcile inconsistent data sets. Journal of Land use Science, 3(4):
251--264. \\

Wang, F. (1990). Fuzzy supervised classification of remote sensing
images. Geoscience and Remote Sensing, IEEE Transactions on, 28(2),
194-201.\\

Zadeh, L.A., (1965). Fuzzy sets. Information and Control 8, 338--353.\\


